Draft Article

### 2020-04-19 article.tex
Made clean-up edits from dad

\title{Draft Article}
\author{Alexander W. Petros}
\date{September 2019}

@@

In 2015, federal authorities discovered that malicious actors had for months maintained access to databases within the United States Office of Personnel Management (OPM). Among other types of sensitive data, these hackers stole a trove of applications for federal security clearances, securing not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses. The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could jeopardize intelligence efforts by exposing undercover CIA officers, or identify the American officials who might be susceptible to blackmail, or worse, recruitment. \enquote{if you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}. In the case of the OPM hack, imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}. But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences. To the extent that phenomenon has been studied politically or academically, it is often explained away as a logical extension of existing policy: cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage. This article seeks to unravel that tautology by explaining how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction serves, and whether it is prudent to continue making it. The transfer of espionage into the digital realm has made that task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm. ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack. \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As far as the public knows, however, the size of operation did \emph{not} change the nature the diplomatic response. Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.

@@

To evaluate each state's incentives I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for how a tacit accord permitting general espionage might be mutually strategic.

@@

\subsection{Espionage norms}
There are few extant definitions of peacetime espionage, and no satisfying ones. The problem that aspiring scholars of espionage inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost entirely missing from any legal codes or case law. Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't attempt to \enquote{define} espionage as such, but rather offer a working definition that bounds the scope of the paper. See \cite[p.~600]{radsan_unresolved_2007},\cite[p.~2]{kapp_spying_2007})} Excluding this activities is important; because assassinations and similar covert operations are not considered espionage they do not typically receive the protections of espionage norms.

For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage. The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory. In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against them, the consequences imposed on the spier remain minimal. Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them. In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.

Once again, we can use cyber-espionage as an analogue for traditional espionage writ large. Though not codified in US policy, the informal regime of espionage norms is an open secret. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets. See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response. CIA Director John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}

@@

\subsection{Existing literature}
Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim determines that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

>>>>

\title{Draft Article}
\author{Alexander W. Petros}
\date{April 2020}

@@

In 2015, federal authorities discovered that malicious actors had for months maintained access to databases within the United States Office of Personnel Management (OPM). The hackers downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses. The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, or identify American officials who might be susceptible to blackmail, or worse, recruitment. \enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}. In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}. But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences. Within both political and academic circles, this phenomenon is often rationalized as a logical extension of existing policy; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage. This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves. The practice of espionage in the digital realm has made this task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm, and governments re-evaluate whether to employ decades-old norms. Nations like Brazil are openly pushing for espionage to be regulated through larger regimes of Internet governance.\footcite[p.~464]{abdenur_triggering_2015} ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'', said one anonymous senior intelligence official of the OPM hack. \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} For now, however, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.

% Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.

@@

To evaluate each state's incentives I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for a mutually strategic tacit accord permitting peacetime espionage.

@@

\subsection{Espionage norms}
There are few extant definitions of peacetime espionage, and no satisfying ones. The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but is almost entirely missing from any legal codes or case law. Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper. See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important; because assassinations and similar covert operations are not considered espionage, they do not typically receive the protections of espionage norms.

For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage. The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory. In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against it, the consequences imposed on the spying country remain minimal. Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them. In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.

Once again, we can use cyber-espionage as an analogue for traditional espionage. Though not codified in US policy, the informal regime of espionage norms is an open secret. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets. See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response. CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM]. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}

@@

\subsection{Existing literature}
Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system happily leaves this tension in a state of salutary neglect. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.
